Critical assessment of Linear-Response Excitations in Molecular system\\ or \\ On the Localization of the Linear-Response excitations in Open Systems

### 2018-02-15 LRDraft.tex
update
Update on Overleaf.

\section{Reboot with recent considerations}
\begin{itemize}
\item Write the excitations as the eigenstates of the Liouvillian:
\begin{itemize}
\item The Casida Matrix provides the eigenvectors in the basis of the unperturbed transitions
\item From the Casida eigenproblem we 
get the poles of the dynamical polarizability and we may reconstruct the perturbed wavefunctions
\item Lot of solution in state of 
the art to express \emph{the same information}, still by avoiding the explicit diagonalization of the problem
\end{itemize}
\item Such transitions might also be written in terms of the solutions of the eigenproblem for the perturbed hamiltonian
\begin{itemize}
\item There are therefore excitations which correspond to localized solutions of the above eigenproblems
\item There is a \emph{energy threshold} above which the eigenfunction associated \emph{cannot be anymore} localized
\end{itemize}
\item Below threshold, the eigenfunction are localized and the energy values are \emph{observable} quantities (there is a convergence in terms of the simulation parameters)
\begin{itemize}
\item Casida treatments for different
basis sets exhibit convergence of the low-energy excitations
\item Still, even for such localized excitations, non-trivial dependence on the empty states on some of excitation value is observed (energies converge relatively slowly wrt $N_\alpha$)
\item The locality of the excitation can be \emph{measured}
{\bf NEW idea for the SoB: why not to consider}
$$
\int \mathrm d \mathbf r
\rho(\mathbf r) |\phi_p^E(\mathbf r)|^2
$$
or similar quantities, that are independent on the number of empty and  bound states
\item Localized Excitation are the  only one with an observable character (this is surely already known)
\item The spectrum of the Liouvillian in this regime is a \emph{discrete} spectrum; the excitation energies are \emph{poles} of the dynamical polarizability
nano\item We may associate the value of the threshold to the \emph{ionization potential}: This is a very interesting remark in my opinion: the physics of the optical excitation is observable 
(and therefore localized) only if the energies are not higher than the energy needed to ionize the system. After this point, it seems normal to me that the ``optical'' excitations with 
real value of the energy are not  meant to be observable anymore.
\item Optical Excitations below threshold have therefore a (very) long lifetime (poles on the real axis), and they contribute to the analytic structure of the dynamical polarizability.
\end{itemize}
\item Above this threshold the energies have a strong dependence on the computational treatment
\begin{itemize}
\item The high-energy part of the 
spectrum does not converge with the 
boundary conditions, even though they converge wrt the empty states (established techniques \emph{do not solve} such problem)
\item This is \emph{not} a problem that is related to the continuum collapse of the empty states.
\item They belong to the continuum of  the Liouvillian $\rightarrow$ they collapse, or depend strongly on the basis set adopted
\item They cannot be considered separately \emph{even when they are expressed in localized basis} as 
their shape is dependent on the treatment
\item They energies are not \emph{true} poles of the dynamical polarizability (the spectra converge only by superposition of deltas)
\end{itemize}
\item Optical excitation above Ionization threshold are therefore observable \emph{only} if they possess a lifetime (the peaks of the spectrum have a well-defined broadening). 
In this way they might be expressed in terms of resonant wavefunctions
\end{itemize}

@@

Much more interestingly for the main topic of the present discussion, and regardless from the detailed strategy of their solutions, equations \eqref{fluctuationStateEqMotion1} 
allow us to present a \emph{localization argument} for the fluctuation states. We discuss this point for the right vectors, \emph{i.e.} for the first line of \eqref{fluctuationStateEqMotion1}, 
but the same argument holds also for the left ones. We split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term, so that the equation of motion 
for $\ket{f_p(\omega)}$ is recasted as:
\be\lb{fluctuationStateEqMotion}
\left(\bbomega+\eps_p -\op T\right)\ket{f_p(\omega)} = \op V\ket{f_p(\omega)} + \op Q_0\left(\op V'(\omega) + \op \Phi(\omega)\right)\ket{\psi_p} \;.
\ee
This equation has the formal structure of an inhomogeneous Helmholtz equation and can be written in the coordinate representation as:
\be
\left(k^2_p(\omega)+\ii\eta + \nabla^2\right)f_p(\omega,\r) = s(\omega,\r) \;\;, {\rm with} \;\; k^2_p(\omega) = 2(\omega+\eps_p) \;.
\ee
The solution of this equation admits an integral representation in the form:
\be\lb{fluctuationStateIntEq1}
f_p(\omega,\r) = \int \dd\r' g_H(\r-\r')s(\r') \;,
\ee
where $g_H(\r)$ is the Green function of the Helmholtz operator in three dimensions. This integral representation allows us to assess the localization character 
of the fluctuation states, \emph{i.e.} the behavior of $f_p(\omega,\r)$ when $\r$ becomes large. Indeed, assuming that the support of the source term is compact
we have that, for $\r$ outside the domain in which the integrand of \eqref{fluctuationStateIntEq1} is not vanishing, the leading contribution of the fluctuation states
is given by:
\be\lb{fluctuationStateIntEq2}
f_p(\omega,\r) \sim \frac{e^{\ii \sqrt{k^2_p(\omega)+\ii\eta} |\r|}}{|\r|} \;.
\ee
We observe that the presence of the regularization parameter $\eta$ ensures a (slow) asymptotic convergence to zero, nonetheless the qualitative $\r$-dependence of the fluctuation 
states is determined by the sign of $k^2_p(\omega)$. In particular equation \eqref{fluctuationStateIntEq2} describes an exponential damping for $\omega < |\eps_p|$ and an oscillating 
power law behavior for $\omega \geq |\eps_p|$.
This argument holds for all the fluctuation states that contribute to the perturbation-induced expectation value \eqref{LinearResponseFunctDef3}, so we conclude that
the linear response functional is built in terms of localized fluctuation states for $\omega < |\eps_h|$. Instead, in the range $\eps_h|\leq\omega<|\eps_0|$ it contains both
contributions from local and non-local quantities and, finally is composed only by non-local fluctuation states for $\omega\geq|\eps_0|$.

@@

We perform a particular choice of the perturbing field:
\be
\op\Phi = \commutator{\op E}{\dmnot} \; \longrightarrow \; \opskew{\Phi} = 
\commutator{\commutator{\op E}{\dmnot}}{\dmnot} = \op E
\ee

>>>>

\section{Reboot with recent considerations}
Given a approximate eigenvector $\ket{\psi}$ we approximate its energy by $E=\bra{\psi} \op H \ket{\psi}$.
This energy is an approximation of the true eigenvalue. We assume that the wavefunction gradient $\ket{g(E)}\equiv \left(\op H -E\right)\ket{\psi}$ indicates the distance from the solution of the Schroedinger Equation (SE) in a complete basis set.
By imposing the solution of the SE for a better approximation of the wavefunction $\psi + \Delta \psi$
we should obtain, self-consistently:
\begin{align}
\left(E -\op H \right)\ket{\Delta \psi} &= \ket{g(E+ \Delta E)}\;,\\
\Delta E &= 2 \bra{\psi} \op H \ket{\Delta \psi}\;.
\end{align}
From the expression of $\Delta \psi$ we can recalculate another approximation for the wavefunction and, correspondingly, for the eigenvalue.
By definition, the exact solution is reached only when $\ket{g}=0$, which implies that the norm of $\Delta \psi$ must decrease during the perturbative process.

\textbf{Theorem:} Only the eigenstates with negative energy might have a finite-size correction that can be expressed perturbatvely.

\textbf{Proof}: 
If we apply this process to estimate the finite-size correction to a numerical solution of the Schroedinger equation, as the eigenvalue should converge faster than the eigenvector, we conclude that the gradient $g$ becomes nonzero only in regions that  should be located far from the potential region. If this is true, at the next iteration the gradient would become $g + \op H \Delta \psi + \mathcal{O}(\Delta E)$.
Therefore, for having a convergent perturbative process, also the ket $\ket{\Delta \psi}$ solution of the above equation should be localized far from the support of the potential. In other terms, the relation between $g$ and $\Delta \psi$ should be local in space.

The Lippmann-Schwinger relation defines the Dyson series of the resolvent of the Hamiltonian in terms of the Helmholtz Green's function $\op G_0(E)$:
\op G(E) \equiv (E- \op H)^{-1} = \op G_0(E) + \op G_0(E) \op V \op G(E)\;.
We have therefore 
\ket{\Delta \psi} = 
\sum_{p=0}^\infty \op G_0(E) 
\left(\op V \op G_0(E)\right)^p \ket{g}\;.
For the properties of the Helmholtz equation, locality in space of $\Delta \psi$ might only be recovered for negative energy values. This equation proves that for positive values of the energy the finite-size corrections will \emph{always} provide fully delocalized corrections: it would be impossible to observe a convergence of the wavefunction in the interior region by simply increasing the size of the simulation domain. 
If the value of the approximate energy does not become negative during the perturbative process, 
a modification of the domain boundaries will impact the wavefunction in the entire simulation region. In this case, convergence might only be achieved uniformly and there is no guarantee that a larger domain will provide similar energies.
$\square$
\begin{itemize}
\item Write the excitations as the eigenstates of the Liouvillian:
\begin{itemize}
\item The Casida Matrix provides the eigenvectors in the basis of the unperturbed transitions
\item From the Casida eigenproblem we 
get the poles of the dynamical polarizability and we may reconstruct the perturbed wavefunctions
\item Lot of solution in state of 
the art to express \emph{the same information}, still by avoiding the explicit diagonalization of the problem
\end{itemize}
\item Such transitions might also be written in terms of the solutions of the eigenproblem for the perturbed hamiltonian
\begin{itemize}
\item There are therefore excitations which correspond to localized solutions of the above eigenproblems
\item There is a \emph{energy threshold} above which the eigenfunction associated \emph{cannot be anymore} localized
\end{itemize}
\item Below threshold, the eigenfunction are localized and the energy values are \emph{observable} quantities (there is a convergence in terms of the simulation parameters)
\begin{itemize}
\item Casida treatments for different
basis sets exhibit convergence of the low-energy excitations
\item Still, even for such localized excitations, non-trivial dependence on the empty states on some of excitation value is observed (energies converge relatively slowly wrt $N_\alpha$)
\item The locality of the excitation can be \emph{measured}
{\bf NEW idea for the SoB: why not to consider}
$$
\int \mathrm d \mathbf r
\rho(\mathbf r) |\phi_p^E(\mathbf r)|^2
$$
or similar quantities, that are independent on the number of empty and  bound states
\item Localized Excitation are the  only one with an observable character (this is surely already known)
\item The spectrum of the Liouvillian in this regime is a \emph{discrete} spectrum; the excitation energies are \emph{poles} of the dynamical polarizability
nano\item We may associate the value of the threshold to the \emph{ionization potential}: This is a very interesting remark in my opinion: the physics of the optical excitation is observable 
(and therefore localized) only if the energies are not higher than the energy needed to ionize the system. After this point, it seems normal to me that the ``optical'' excitations with 
real value of the energy are not  meant to be observable anymore.
\item Optical Excitations below threshold have therefore a (very) long lifetime (poles on the real axis), and they contribute to the analytic structure of the dynamical polarizability.
\end{itemize}
\item Above this threshold the energies have a strong dependence on the computational treatment
\begin{itemize}
\item The high-energy part of the 
spectrum does not converge with the 
boundary conditions, even though they converge wrt the empty states (established techniques \emph{do not solve} such problem)
\item This is \emph{not} a problem that is related to the continuum collapse of the empty states.
\item They belong to the continuum of  the Liouvillian $\rightarrow$ they collapse, or depend strongly on the basis set adopted
\item They cannot be considered separately \emph{even when they are expressed in localized basis} as 
their shape is dependent on the treatment
\item They energies are not \emph{true} poles of the dynamical polarizability (the spectra converge only by superposition of deltas)
\end{itemize}
\item Optical excitation above Ionization threshold are therefore observable \emph{only} if they possess a lifetime (the peaks of the spectrum have a well-defined broadening). 
In this way they might be expressed in terms of resonant wavefunctions
\end{itemize}

@@

Much more interestingly for the main topic of the present discussion, and regardless from the detailed strategy of their solutions, equations \eqref{fluctuationStateEqMotion1} 
allow us to present a \emph{localization argument} for the fluctuation states. We discuss this point for the right vectors, \emph{i.e.} for the first line of \eqref{fluctuationStateEqMotion1}, 
but the same argument holds also for the left ones. We split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term, so that the equation of motion 
for $\ket{f_p(\omega)}$ is recasted as:
\be\lb{fluctuationStateEqMotion}
\left(\bbomega+\eps_p -\op T\right)\ket{f_p(\omega)} = \op V\ket{f_p(\omega)} + \op Q_0\left(\op V'(\omega) + \op \Phi(\omega)\right)\ket{\psi_p} \;.
\ee
This equation has the formal structure of an inhomogeneous Helmholtz equation and can be written in the coordinate representation as:
\be
\left(k^2_p(\omega)+\ii\eta + \nabla^2\right)f_p(\omega,\r) = s(\omega,\r) \;\;, {\rm with} \;\; k^2_p(\omega) = 2(\omega+\eps_p) \;.
\ee
The solution of this equation admits an integral representation in the form:
\be\lb{fluctuationStateIntEq1}
f_p(\omega,\r) = \int \dd\r' g_H(\r-\r')s(\r') \;,
\ee
where $g_H(\r)$ is the Green function of the Helmholtz operator in three dimensions. This integral representation allows us to assess the localization character 
of the fluctuation states, \emph{i.e.} the behavior of $f_p(\omega,\r)$ when $\r$ becomes large. Indeed, assuming that the support of the source term is compact
we have that, for $\r$ outside the domain in which the integrand of \eqref{fluctuationStateIntEq1} is not vanishing, the leading contribution of the fluctuation states
is given by:
\be\lb{fluctuationStateIntEq2}
f_p(\omega,\r) \sim \frac{e^{\ii \sqrt{k^2_p(\omega)+\ii\eta} |\r|}}{|\r|} \simeq 
e^{-\frac{\eta}{2}|\r|} \frac{e^{\ii \sqrt{k^2_p(\omega)} |\r|}}{|\r|} \;.
\ee
We observe that the presence of the regularization parameter $\eta$ ensures a (slow) asymptotic convergence to zero, nonetheless the qualitative $\r$-dependence of the fluctuation 
states is determined by the sign of $k^2_p(\omega)$. In particular equation \eqref{fluctuationStateIntEq2} describes an exponential damping for $\omega < |\eps_p|$ and an oscillating 
power law behavior for $\omega \geq |\eps_p|$.
This argument holds for all the fluctuation states that contribute to the perturbation-induced expectation value \eqref{LinearResponseFunctDef3}, so we conclude that
the linear response functional is built in terms of localized fluctuation states for $\omega < |\eps_h|$. Instead, in the range $\eps_h|\leq\omega<|\eps_0|$ it contains both
contributions from local and non-local quantities and, finally is composed only by non-local fluctuation states for $\omega\geq|\eps_0|$.

@@

We perform a particular choice of the perturbing field:
\be
\op\Phi = \commutator{\op E}{\dmnot} \; \longrightarrow \; \opskew{\Phi} = 
\commutator{\commutator{\op E}{\dmnot}}{\dmnot} = \op E
\ee
...to be continued...we have to state precisely the time-dependence of $\hone{t}$...
